Joint estimation of phylogenies and ancestral states is not consistent

### 2017-12-22 appendix.tex
napkin math in the comments
moved comments to words

%EM I think this following statement could use a little more justification. If we write out a symmetric function f in two variables and take the derivative of f(x+delta, x) = f(x, x+delta) with respect to delta and set it to zero we can see that this must be a stationary point of f. Do we need to say more to ensure that this is a maximum?
%das: I wasn't even thinking derivatives. Let f(x, y) be symmetric so that f(x, y) = f(y, x). If f is maximized at some (x_0, y_0) then f(x_0, y_0) > f(x, y) for all x, y. But then by symmetry f(y_0, x_0) is also a maximum. Now *if* a unique maximum exists, x_0 = y_0 should necessarily be this maximum by the properties defined above, otherwise, f(x_0, y_0) would not be equal to f(y_0, x_0). I think there are probably counterexamples where there is not a unique maximum where this statement isn't necessarily true (trivially, because you'd just say x_0 = y_1 if y_1 is also a maximum and voila) but I don't think this discounts the fact that x_0 = y_0 will still be *a* maximum in this case. In that case, it's reasonable to assume so, in my opinion, especially for the topological inconsistency (we don't use the branch length values), but even for the branch length inconsistency because the case of no unique solution is just as bad if not worse than inconsistency.
%das: Should some or all of that be in the text? Can you see anything wrong with it?
%EM I don't see why this works. In fact, imagine that x=y is a ridge of minima, and that we have maxima at (1,2) and at (2,1). My argument shows that we must have a stationary point at x=y, but that doesn't mean that it has to be a maximum!
%das: Wouldn't the assumption of there existing a unique maximum preclude your case? My argument is that if there is a unique maximum, it necessarily has to have x=y. If there is not a unique maximum then the ML problem is ill-posed. I now don't see why we have to have a unique maximum other than to say we hope the problem is well-posed, but it may not be. Do you have any thoughts on where to take this?
%das: So there are references for identifiability, but I'm a little nervous using them because they're for the usual marginal ML problem. How cagey does it sound to say (as I do below) "here are references for identifiability in the usual case, so either joint inf will be not identifiable (bad) or it (likely?) will be and we can use this (which turns out to be inconsistent later)?" Can you think of a better approach?
%das: some other references I found for identifiability are Rogers 1997, RoyChoudhury 2014...
The maximum likelihood procedure in phylogenetics results in identifiable parameter estimation under mild conditions \cite{Chang1996}.
The joint inference procedure may not necessarily result in the same identifiability properties, but we claim the following.
If joint inference obtains unique parameter estimates, then, since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, $x_1=x_2$ at the maximum, with the analogous statement holding for $y_1$ and $y_2$.
The case in which we do not obtain unique parameter estimates results in an ill-posed likelihood problem, and thus by design an inconsistency since multiple maxima exist that will not equal the true, generating parameters.
Given this, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property of symmetry of function inputs, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{w-hat-empirical-01-bias}
\caption{
    Estimates for $\hat{w}-y^*$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:marginal_likelihood}.
    Plot focuses on $0.6 < x^*, y^* < 1$ where $\hat{w}$ is estimated to be a value different from 1.
}
\label{fig:bl-general-bias}
\end{figure}

>>>>

%EM I think this following statement could use a little more justification. If we write out a symmetric function f in two variables and take the derivative of f(x+delta, x) = f(x, x+delta) with respect to delta and set it to zero we can see that this must be a stationary point of f. Do we need to say more to ensure that this is a maximum?
%das: I wasn't even thinking derivatives. Let f(x, y) be symmetric so that f(x, y) = f(y, x). If f is maximized at some (x_0, y_0) then f(x_0, y_0) > f(x, y) for all x, y. But then by symmetry f(y_0, x_0) is also a maximum. Now *if* a unique maximum exists, x_0 = y_0 should necessarily be this maximum by the properties defined above, otherwise, f(x_0, y_0) would not be equal to f(y_0, x_0). I think there are probably counterexamples where there is not a unique maximum where this statement isn't necessarily true (trivially, because you'd just say x_0 = y_1 if y_1 is also a maximum and voila) but I don't think this discounts the fact that x_0 = y_0 will still be *a* maximum in this case. In that case, it's reasonable to assume so, in my opinion, especially for the topological inconsistency (we don't use the branch length values), but even for the branch length inconsistency because the case of no unique solution is just as bad if not worse than inconsistency.
%das: Should some or all of that be in the text? Can you see anything wrong with it?
%EM I don't see why this works. In fact, imagine that x=y is a ridge of minima, and that we have maxima at (1,2) and at (2,1). My argument shows that we must have a stationary point at x=y, but that doesn't mean that it has to be a maximum!
%das: Wouldn't the assumption of there existing a unique maximum preclude your case? My argument is that if there is a unique maximum, it necessarily has to have x=y. If there is not a unique maximum then the ML problem is ill-posed. I now don't see why we have to have a unique maximum other than to say we hope the problem is well-posed, but it may not be. Do you have any thoughts on where to take this?
%das: So there are references for identifiability, but I'm a little nervous using them because they're for the usual marginal ML problem. How cagey does it sound to say (as I do below) "here are references for identifiability in the usual case, so either joint inf will be not identifiable (bad) or it (likely?) will be and we can use this (which turns out to be inconsistent later)?" Can you think of a better approach?
%das: some other references I found for identifiability are Rogers 1997, RoyChoudhury 2014...
Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, $x_1=x_2$ at the maximum provided a unique maximum exists.
An analogous statement holds for $y_1$ and $y_2$.
It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference, similar arguments holding for the remaining parameters.
We decompose the likelihood into the entropy and partial likelihood terms as in \eqref{eq:log_likelihood_simplified}.
By Gibbs's inequality, the entropy term $\shannonDivergence_{\tau^*,t^*}(\tau^*,t)$ has a unique maximum over $t$---namely $t^*$.
For the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$, we can compute the partial likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ as in Table~\ref{tab:farris_likelihoods}.
Clearly this partial likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $1/2 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in[0,1]$ and thus has a unique maximum.
Since both summands of the likelihood are concave on $x_1\in[0,1]$, the likelihood is as well, and there exists a single $x_1\in[0,1]$ maximizing the likelihood.
Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property of symmetry of function inputs, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{w-hat-empirical-01-bias}
\caption{
    Estimates for $\hat{w}-y^*$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:marginal_likelihood}.
    Plot focuses on $0.6 < x^*, y^* < 1$ where $\hat{w}$ is estimated to be a value different from 1.
    We do not compute the bias for the white region where $\hat{w}=1$.
}
\label{fig:bl-general-bias}
\end{figure}
